---
title: "Coronavirus data visualisation"
output: 
  html_document:
   toc: true
   toc_float: true
   logo: LogoAgrocampusOuest.jpg
   number_sections: false
   highlight: tango
   css: style.css
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, cache = TRUE)
```

The [Johns Hopkins University](https://coronavirus.jhu.edu/) provide data on the coronavirus crisis, reportin the daily number of confirmed cases,  deaths and recovered patients.



## Data processing 


### Download data 

```{r load_data}
library('readr')
library('tidyverse')


url_base <- 
  "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_19-covid"

type <- c('Confirmed', 'Deaths', 'Recovered')
dta_list <- lapply( type, function(type_){
 read_csv(url(paste0(url_base, '-', type_, '.csv'))) %>% 
    mutate(type = type_)
})

dta <- do.call('rbind', dta_list)
```


### Formatting data

```{r format_data}
dta %>% 
  rename(Province = `Province/State`, Country = `Country/Region`) %>% 
  pivot_longer(-c(Province, Country, Lat, Long, type),  names_to = 'Date', values_to = 'Cases') -> dta

dta %>%  
   mutate(Date = lubridate::mdy(Date))  %>% 
   mutate(country_id = tolower(Country)) %>% 
   group_by(country_id, type, Date) %>% 
   summarise(Cases = sum(Cases, na.rm = TRUE)) %>% 
   ungroup()-> dta
```



## Spatial Spreading of the pandemy 

```{r  echo = FALSE}
n_tot <- n_distinct(dta$country_id)
n1 <- 500
n2 <- 8000
```

Data are available from `r min(dta$Date)` to `r max(dta$Date)` for `r n_tot` countries.

### Define the severity of the pandemy in every country

The visualisation is done with only the most affected countries. The severity of the pandemy in a country is said to 
- Low if there are less than `r n1`  confirmed cases
- High if there are at least  `r n1+1`  confirmed cases
- Very High low if there are at least `r n2+1`  confirmed cases



```{r}

dta %>%   filter(type == 'Confirmed') %>% group_by(country_id) %>%
   summarise ( m = max(Cases, na.rm = TRUE)) %>%  filter( m  > n1) -> country_list

dta %>%   filter(type == 'Confirmed') %>% group_by(country_id) %>%
   summarise ( m = max(Cases, na.rm = TRUE)) %>%  filter( m  > n2) -> country_list_vhigh

dta %>% filter(country_id %in% country_list$country_id) -> dta


severity_class <- function(country){
  prov <- ifelse(country %in% country_list$country_id, 'High', 'Low')
  prov <-  ifelse(country %in% country_list_vhigh$country_id, 'Very High', prov)
  prov <- factor(prov, levels = c('Low', 'High', 'Very High'))
  return(prov)
}

dta %>% mutate(Severity = severity_class(country_id)) -> dta
```




### Mapping the data

The `sf` library, allows to produce map easily. 

Mapping the  data on a world map requires two merge the world shapefile and the Johns Hoppkins data as proposed in the following code. The matching between the two databases  is done by creating a  `country_id` identification (country name in lower case). The few mismatches in the 2 databases are idenified and corrected. 


```{r}
library(sf)

world_map <- st_read(dsn = 'datasets/shape_dir/TM_WORLD_BORDERS-0.3.shp') %>% 
  st_as_sf() %>% mutate(country_id = tolower(NAME))

country_list %>% filter( !country_id %in% world_map$country_id) -> country_issues
country_issues$country_id

world_map %>% mutate(avail = ifelse(country_id %in% country_list$country_id, TRUE, FALSE)) -> world_map

world_map %>% 
  mutate( country_id = ifelse(country_id == "united states", 'us' , country_id))%>% 
  mutate( country_id = ifelse(country_id == "iran (islamic republic of)", 'iran' , country_id)) %>% 
  mutate( country_id = ifelse(country_id == "korea, republic of", "korea, south" , country_id)) %>% 
  mutate( country_id = ifelse(country_id == "czech republic", "czechia" , country_id)) -> world_map 
```


 Finally the three levels of severity are  presented in the map below

```{r}
world_map %>%  mutate(Severity = severity_class(country_id)) -> world_map

world_map %>% ggplot() + geom_sf(aes(fill = Severity)) + 
  scale_fill_viridis_d()  
```



## Time series data



### Confirmed cases


We can represent the evolution of coronavirus confirmed cases for the countries with high severity. 

```{r}
dta %>% filter(type == 'Confirmed',  Severity != 'Low') %>% ggplot() + geom_line(aes(x= Date, y = Cases, col = country_id)) + scale_y_continuous(trans='log2') + theme(legend.title = element_text( size = 10),
  legend.text = element_text(size = 6) ) + ggtitle('Evolution of confirmed cases in highly affected countries')
```


Too many data on the graph , focus on the countries with High level of severity.

```{r echo = FALSE}
couleurs <- c('#ff0000', '#ff9900', '#006600', 
              '#2d5986', '#600080', '#b30000',
              '#00cc00', '#ff8c66', '#cc0052', 
              '#3939ac')
```


```{r, fig.cap='Evolution of the number of confirmed cases'}

p <- dta  %>% 
   filter(type == 'Confirmed', Severity == 'Very High') %>% 
   arrange(country_id, Date) %>% 
   group_by(country_id) %>%  
   ggplot() + 
   geom_line(aes(x= Date, y = Cases, col = country_id))  + 
   scale_color_manual(values = couleurs) 
p
```



and with a log scale 

```{r, fig.cap='Evolution of the number of confirmed cases in log2 scale'}
p +  
   scale_y_continuous(trans='log2')
```


**Change time origine** :  Accounting for the delay in the spread of the pandemy, we could decide to represent time since the first instant where the 80$^{th}$ cases were confirmed.



Presented below is the number of confirmed cases. 

```{r, fig.cap='Evolution of confirmed cases, time elapsed since the 80th reported cases'}
dta %>% 
   filter(type == 'Confirmed', Cases >= 80,Severity == 'Very High') %>% 
   group_by(country_id) %>% 
   arrange(Date) %>% 
   mutate(Date = Date - dplyr::first(Date)) %>% 
   ggplot() + geom_line(aes(x= Date, y = Cases, col = country_id))  +
   scale_color_manual(values = couleurs)  + 
   scale_y_continuous(trans='log2') 
```

### Evolution of the number of deaths



Presented below is the number of deaths. Those time series are more reliable, as they do not depend on the test effort.

```{r, fig.cap='Evolution of the number of deaths, time elapsed since the 80th reported cases'}
dta %>% 
   pivot_wider(names_from = type, values_from = Cases) %>% 
   filter( Confirmed >= 80,Severity == 'Very High') %>% 
   group_by(country_id) %>% 
   arrange(Date) %>% 
   mutate(Date = Date - dplyr::first(Date)) %>% 
   ggplot() + 
   geom_line(aes(x= Date, y = Deaths, col = country_id)) +
   scale_color_manual(values = couleurs) +
   scale_y_continuous(trans='log2')
```


As the number of confirmed cases might not be the best indicator of the spread of the pandemy, we choose, time of the tenth  as time origine.


```{r, fig.cap='Evolution of the number of deaths, time elapsed since the 10th death'}
dta %>% 
   pivot_wider(names_from = type, values_from = Cases) %>% 
   filter( Deaths >= 10,Severity == 'Very High') %>% 
   group_by(country_id) %>% 
   arrange(Date) %>% 
   mutate(Date = Date - dplyr::first(Date)) %>% 
   ggplot() + geom_line(aes(x= Date, y = Deaths, col = country_id))  +
   scale_color_manual(values = couleurs)  +
   scale_y_continuous(trans='log2') 
```

