---
title: "Coronavirus data visualisation"
output: 
  html_document:
   toc: true
   toc_float: true
   logo: LogoAgrocampusOuest.jpg
   number_sections: false
   highlight: tango
   css: style.css
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

The [Johns Hopkins University](https://coronavirus.jhu.edu/) provide data on the coronavirus crisis, reportin the daily number of confirmed cases,  deaths and recovered patients.



## Data processing 


### Download data 

```{r load_data}
library('readr')
library('tidyverse')


url_base <- 
  "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_"

type <- c('confirmed', 'deaths', 'recovered')
dta_list <- lapply( type, function(type_){
   dta_prov <- read_csv(url(paste0(url_base,  type_, '_global.csv')))  %>% 
      mutate(type = type_) 
   dta_prov %>% 
      rename(Province = `Province/State`, Country = `Country/Region`) %>% 
      pivot_longer(-c(Province, Country, Lat, Long, type),  names_to = 'Date', values_to = 'Cases')  %>% 
      mutate(Date = lubridate::mdy(Date))
})

dta <- do.call('rbind', dta_list)
```

```{r  echo = FALSE}
n_tot <- n_distinct(dta$Country)
n1 <- 500
n2 <- 10000
```

**Data are available from `r min(dta$Date)` to `r max(dta$Date)` for `r n_tot` countries.**


### Formatting data

```{r format_data}


dta %>%  
   mutate(type = factor(type, levels = c('confirmed', 'deaths', 'recovered')))  %>% 
   mutate(country_id = factor(tolower(Country))) %>% 
   group_by(country_id, type, Date) %>% 
   summarise(Cases = sum(Cases, na.rm = TRUE)) %>% 
   ungroup()-> dta
```



## Spatial Spreading of the pandemy 

```{r  echo = FALSE}
n_tot <- n_distinct(dta$country_id)
n1 <- 500
n2 <- 10000
```

Data are available from `r min(dta$Date)` to `r max(dta$Date)` for `r n_tot` countries.

### Define the severity of the pandemy in every country

The visualisation is done with only the most affected countries. The severity of the pandemy in a country is said to 
- Low if there are less than `r n1`  confirmed cases
- High if there are at least  `r n1+1`  confirmed cases
- Very High low if there are at least `r n2+1`  confirmed cases



```{r}

dta %>%   filter(type == 'confirmed') %>% group_by(country_id) %>%
   summarise ( m = max(Cases, na.rm = TRUE)) %>%  filter( m  > n1) -> country_list

dta %>%   filter(type == 'confirmed') %>% group_by(country_id) %>%
   summarise ( m = max(Cases, na.rm = TRUE)) %>%  filter( m  > n2) -> country_list_vhigh

dta %>% filter(country_id %in% country_list$country_id) -> dta


severity_class <- function(country){
  prov <- ifelse(country %in% country_list$country_id, 'High', 'Low')
  prov <-  ifelse(country %in% country_list_vhigh$country_id, 'Very High', prov)
  prov <- factor(prov, levels = c('Low', 'High', 'Very High'))
  return(prov)
}

dta %>% mutate(Severity = severity_class(country_id)) -> dta
```




### Mapping the data

The `sf` library, allows to produce map easily. 

Mapping the  data on a world map requires two merge the world shapefile and the Johns Hoppkins data as proposed in the following code. The matching between the two databases  is done by creating a  `country_id` identification (country name in lower case). The few mismatches in the 2 databases are idenified and corrected. 


```{r}
library(sf)

world_map <- st_read(dsn = 'datasets/shape_dir/TM_WORLD_BORDERS-0.3.shp') %>% 
  st_as_sf() %>% mutate(country_id = tolower(NAME))

country_list %>% filter( !country_id %in% world_map$country_id) -> country_issues
country_issues$country_id

world_map %>% mutate(avail = ifelse(country_id %in% country_list$country_id, TRUE, FALSE)) -> world_map

world_map %>% 
  mutate( country_id = ifelse(country_id == "united states", 'us' , country_id))%>% 
  mutate( country_id = ifelse(country_id == "iran (islamic republic of)", 'iran' , country_id)) %>% 
  mutate( country_id = ifelse(country_id == "korea, republic of", "korea, south" , country_id)) %>% 
  mutate( country_id = ifelse(country_id == "czech republic", "czechia" , country_id)) -> world_map 
```


 Finally the three levels of severity are  presented in the map below

```{r}
world_map %>%  mutate(Severity = severity_class(country_id)) -> world_map

world_map %>% ggplot() + geom_sf(aes(fill = Severity)) + 
  scale_fill_viridis_d()  
```



## Time series data



### confirmed cases


We can represent the evolution of coronavirus confirmed cases for the countries with high severity. 

```{r}
dta %>% filter(type == 'confirmed',  Severity != 'Low') %>% ggplot() + geom_line(aes(x= Date, y = Cases, col = country_id)) + scale_y_continuous(trans='log2') + theme(legend.title = element_text( size = 10),
  legend.text = element_text(size = 6) ) + ggtitle('Evolution of confirmed cases in highly affected countries')
```


Too many data on the graph , focus on the countries with High level of severity.

```{r echo = FALSE}
couleurs <- c('#ff0000', '#ff9900', '#006600', 
              '#2d5986', '#600080', '#b30000',
              '#00cc00', '#ff8c66', '#cc0052', 
              '#3939ac', '#a5fc03', '#ff0000',
              '#ab0a9b')
```


```{r, fig.cap='Evolution of the number of confirmed cases'}

p <- dta  %>% 
   filter(type == 'confirmed', Severity == 'Very High') %>% 
   arrange(country_id, Date) %>% 
   group_by(country_id) %>%  
   ggplot() + 
   geom_line(aes(x= Date, y = Cases, linetype=country_id, col = country_id))  + 
   scale_color_manual(values = couleurs) 
p
```



and with a log scale 

```{r, fig.cap='Evolution of the number of confirmed cases in log2 scale'}
p +  
   scale_y_continuous(trans='log2')
```


As the pandemy does not start to spread at the same date on every country, one needs to define some new origine of time for every country, in order to allow the comparison between countries.


**Change time origine** :  Accounting for the delay in the spread of the pandemy, we could decide to represent time since the first instant where the 80$^\mbox{th}$ cases were confirmed.



Presented below is the number of confirmed cases. 

```{r, fig.cap='Evolution of confirmed cases, time elapsed since the 80th reported cases'}
dta %>% 
   filter(type == 'confirmed', Cases >= 80,Severity == 'Very High') %>% 
   group_by(country_id) %>% 
   arrange(Date) %>% 
   mutate(Date = Date - dplyr::first(Date)) %>% 
   ggplot() + geom_line(aes(x= Date, y = Cases, linetype=country_id, col = country_id, ))  +
   scale_color_manual(values = couleurs)  + 
   scale_y_continuous(trans='log2') 
```

### Evolution of the number of deaths



Presented below is the number of deaths. Those time series are more reliable, as they do not depend on the test effort.

```{r, fig.cap='Evolution of the number of deaths, time elapsed since the 80th reported cases'}
dta %>% 
   pivot_wider(names_from = type, values_from = Cases) %>% 
   filter( confirmed >= 80,Severity == 'Very High') %>% 
   group_by(country_id) %>% 
   arrange(Date) %>% 
   mutate(Date = Date - dplyr::first(Date)) %>% 
   ggplot() + 
   geom_line(aes(x= Date, y = deaths, linetype=country_id, col = country_id)) +
   scale_color_manual(values = couleurs) +
   scale_y_continuous(trans='log2')
```


As the number of confirmed cases might not be the best indicator of the spread of the pandemy, we choose,  the time of the tenth  death as the origin of date.


```{r origine_death, fig.cap='Evolution of the number of deaths, time elapsed since the 10th death'}
dta %>% 
   pivot_wider(names_from = type, values_from = Cases) %>% 
   filter( deaths >= 10,Severity == 'Very High') %>% 
   group_by(country_id) %>% 
   arrange(Date) %>% 
   mutate(Date = Date - dplyr::first(Date)) %>% 
   ggplot() + geom_line(aes(x= Date, y = deaths, linetype=country_id, col = country_id))  +
   scale_color_manual(values = couleurs)  +
   scale_y_continuous(trans='log2') 
```

### Evolution of the number of new deaths

As the pandemy is spreading, the number of deaths is expected to grow bigger every day before slowing down. Therefore the number of new daths, which also the derivative of the number of deaths,  could be a good indicator of the stage of the pandemy.


```{r new_death, fig.cap='Evolution of the number of new deaths, time elapsed since the 10th death, log2 scale'}
dta %>% 
   pivot_wider(names_from = type, values_from = Cases) %>% 
   filter( deaths >= 10,Severity == 'Very High') %>% 
   group_by(country_id) %>% 
   arrange(Date) %>% 
   mutate(Date = Date - dplyr::first(Date),
          new_deaths = deaths -lag(deaths)) %>% 
   ggplot() + geom_point(aes(x= Date, y = new_deaths, col = country_id), alpha = 0.3)  +
   geom_smooth(aes(x= Date, y = new_deaths, linetype=country_id, col = country_id), se = FALSE)  +
   scale_color_manual(values = couleurs) +
   scale_y_continuous(trans='log2') 
```





